---
title: "Demographic parameters estimates"
author: "Mellina Sidous"
date: "2026-09-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#load packages
library(readxl)
library(tidyverse)
library(ggplot2)

# load data
data <- read_excel("dat/small_transect_date_corrected.xls") %>% 
  filter(Genero.sp %in% c("Marmosa_murina", "Didelphis_albiventris")) %>% 
  dplyr::select(c(Genero.sp,Localidade,Data_de_coleta,                                                                                    Brinco,codigo, N_de_campo,Faixa_etaria, Sexo, Peso_gramas ,
                  #Embrioes ,Vagina,Grávida, Lactante ,
                  n_Filhotes , reprodutivo ,Habitat_origem, Expedicao,Estacao , 
                  Ocasiao , Estrato_arboreo  )) %>%
  rename(Species = Genero.sp, Fragment = Localidade, Date = Data_de_coleta, Ring = Brinco, ID= N_de_campo, 
         Age = Faixa_etaria, Sex = Sexo, Weight_gramm = Peso_gramas, N_offspring = n_Filhotes, Reproductive_status = reprodutivo, Vegetation_type = Habitat_origem, Sampling_session = Expedicao) %>%
  mutate(Vegetation_type= tolower(Vegetation_type)) %>% 
  mutate (Month =month(data$Date)) %>%  mutate(season = if_else(Month %in% c(2, 3,4,5,6, 7), "humid", "dry"))


```

## How many individuals were resighted ? Summary of the number of individuals captured and recaptured in the different Atlantic forest fragment: 

### First, how many individuals of each species were captured in total ? 

```{r}
unique_individuals <- data %>% group_by(ID) %>% slice(1) 
table(unique_individuals$Species)

```

And how are they spread across the three fragments ?

```{r}
table(unique_individuals$Species, unique_individuals$Fragment)

```
And were they captured within atlantic forest or sugar cane plantation ?

```{r}
table(unique_individuals$Species, unique_individuals$Vegetation_type, unique_individuals$Fragment) 

```
### Among the 305 D. albiventris and 78 M. murina captured in total, how many were captured more than once ?

```{r }
resighted_individuals <- table(data$ID, data$Species) %>% as.data.frame() %>% filter(Freq>1) %>% rename(ID=Var1, Species=Var2, N_captures=Freq)

table(resighted_individuals$Species)

```
127 individuals of Didelphis were captured more than once and 21 Marmosa murina. 

How many times were they resighted?

```{r}
 resighted_individuals %>% group_by(Species) %>% summarise(mean = mean(N_captures), sd = sd(N_captures), max= max(N_captures), min= min(N_captures))

ggplot(resighted_individuals, aes(x= N_captures))+
  geom_histogram()+
  facet_wrap(~Species) +
  theme_bw()


```

In which fragments were they observed (one indiviual count for one, we remove the recapture to obtain the number)?

```{r}
data_resighted_only <- data %>% filter(ID %in% resighted_individuals$ID)

table((data_resighted_only %>% group_by(ID) %>% slice(1))$Species, (data_resighted_only %>% group_by(ID) %>% slice(1))$Fragment)

``` 
### And how many individuals wre recaptured during different sampling sessions ?

```{r}
resighted_different_sampling_sessions <- data_resighted_only %>% group_by(ID, Sampling_session) %>% slice(1)

table((table(resighted_different_sampling_sessions$ID, resighted_different_sampling_sessions$Species) %>% as.data.frame() %>% filter(Freq>1))$Var2) 
```


```{r}
total_resighted_dsampling <- table(resighted_different_sampling_sessions$ID, resighted_different_sampling_sessions$Species) %>% as.data.frame() %>% filter(Freq>1) %>% rename(ID=Var1, Species=Var2, N_captures=Freq)

 total_resighted_dsampling %>% group_by(Species) %>% summarise(mean = mean(N_captures), sd = sd(N_captures), max= max(N_captures), min= min(N_captures))


ggplot(total_resighted_dsampling, aes(x= N_captures))+
  geom_histogram()+
  facet_wrap(~Species) +
  theme_bw()


```

And in what fragments do we have individuals that were recaptured at least once at a different sampling session than that they were captured? 

```{r}

resighted_different_sampling_sessions <- data_resighted_only %>% group_by(ID, Sampling_session) %>% slice(1)

table((data %>% filter( ID %in% total_resighted_dsampling$ID) %>% group_by(ID) %>% slice(1))$Species, (data %>% filter( ID %in% total_resighted_dsampling$ID) %>% group_by(ID) %>% slice(1))$Fragment)


```


## Substract demographic parameters for PVA

### Sex ratio   

One thing we have to be careful is that some individuals were captured several times and thus may count double for these parameters (for example if females are more often recaptured than males, this will bias sex ratio toward females). Thus we have to keep one line per individual


```{r }
data %>% group_by(ID) %>% slice(1) %>% ungroup() %>% filter(Sex %in% c("F", "M")) %>% count(Species, Fragment, Sex) %>% pivot_wider(names_from = Sex, values_from = n) %>% mutate(Prop_females = round(F/(F+M),2))

Sex_ratio <- data %>% group_by(ID) %>% slice(1) %>% ungroup() %>% filter(Sex %in% c("F", "M")) %>% count(Species, Fragment, Sex) %>% pivot_wider(names_from = Sex, values_from = n) %>% mutate(Prop_females = round(F/(F+M),2)) %>% dplyr::select(-c("M", "F")) %>% pivot_wider(names_from = Fragment, values_from = Prop_females ) %>% mutate(Parameter="Sex_ratio")
```

### Percent of reproductive females 

When do the species reproduce ?
Note that we want to focus on adult females. I consider the dry seasons from august to january and the dry from february to july.


```{r}
# per month, consider that the individual is reprodsuctive if it was seen at leats one in a reprodtuve state
#Make sure females appear only once a month
data_monthly <- data %>% filter(Sex == "F") %>% filter(Age == "ADULTO")%>%  group_by(ID, Species, Fragment, Month) %>%
  summarise( Reproductive_status = ifelse( any(Reproductive_status == "S"),"S", "N" ), .groups = "drop" )%>% ungroup()


percent_repro_females <- data_monthly  %>% count(Month, Species, Fragment, Reproductive_status) %>% pivot_wider(names_from = Reproductive_status, values_from = n) %>% mutate(N = ifelse(is.na(N), 0, N), S = ifelse(is.na(S), 0, S)) %>% mutate(Percent_repro = S/(S+N)) %>%  mutate(  Month = factor(Month, levels = 1:12,
      labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun",  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec") ) )



ggplot(percent_repro_females,  aes(x = Month, y = Percent_repro, fill = Fragment)) +
  geom_col( color = "black", position = position_dodge() ) +
  facet_wrap(~Species) +
  scale_fill_manual(values = c( "RPPN_Pacatuba" = "#E69F00", "RPPN_Gargaú" = "darkorchid1", "Rebio_Guaribas" = "yellowgreen" )) +
  theme_bw() +
  labs( x = "Month", y = "Reproductive females (%)",  fill = "Fragment")
```

It seems that Marmosa murina only reproduces in March April (and in Pacatuba), where the proportion of reproductive female is 0.575 (mean between 1 in March and 0.4 in April).

For Didelphis albiventris most reproduction seem to occur during the wet season, from March to July. There is then a slight reproduction during the dry season. Let's see closer the estimates in eah fragment for this species.


```{r}
ggplot((percent_repro_females%>% filter(Species == "Didelphis_albiventris")),  aes(x = Month, y = Percent_repro)) +
  geom_col( color = "black", position = position_dodge() ) +
  facet_wrap(~Fragment) +
  theme_bw() +
  labs( x = "Month", y = "Reproductive females (%)",  fill = "Fragment")


# Proportion by season measured as the mean of the proportion during each given season
percent_repro_females <- percent_repro_females %>% mutate(season = if_else(Month %in%c("Feb", "Mar", "Apr", "May", "Jun", "Jul"), "wet", "dry"))

percent_repro_females %>% group_by(season, Species, Fragment) %>% summarise(mean = mean(Percent_repro))


```


Now calculate this with only one value per female per season

```{r}

data_season <- data %>% filter(Sex == "F")%>% filter(Age == "ADULTO") %>%  group_by(ID, Species, Fragment, season) %>%
  summarise( Reproductive_status = ifelse( any(Reproductive_status == "S"),"S", "N" ), .groups = "drop" )%>% ungroup()


percent_repro_females_season <- data_season  %>% count(season, Species, Fragment, Reproductive_status) %>% pivot_wider(names_from = Reproductive_status, values_from = n) %>% mutate(N = ifelse(is.na(N), 0, N), S = ifelse(is.na(S), 0, S)) %>% mutate(Percent_repro = S/(S+N)) 



final_percent_repro <- percent_repro_females_season %>% dplyr::select(-c(N, S)) %>% pivot_wider(names_from = Fragment, values_from = Percent_repro) %>% mutate(type = "Percent_repro", Parameter = (paste(type, season)%>% str_replace_all(" ", "_"))) %>% dplyr::select(-c(season, type))
# no adult females captured in the dry season in Pacatuba
```

### Age structure 

```{r}
#Do individuals alwaus have the same age within a month ?
data$Year <- year(data$Date)
data <- data %>% filter(!(ID== "MGB_640" & Fragment == "RPPN_Gargaú"))

# First assign only one value when an individual was seen several time: When an individual in the same month is said sub adult and adult, I chose to keep subadult so we only consider them adults when we are sure: If it was seen as reproductive, then it is assigned as an adult

age_monthly <- data %>%
  group_by(ID, Species, Fragment, Year, Month) %>%
  summarise(  Age = case_when(
      any(Age == "JOVEM") ~ "JOVEM",   # Si jovem présent, on garde JOVEM
      
      # Si SUBADULTO et ADULTO sont présents :
      # ADULTO uniquement si reproducteur
      any(Age == "SUBADULTO") & any(Age == "ADULTO") &
        any(Reproductive_status == "S") ~ "ADULTO",
      
      # Sinon, en cas d'hésitation SUBADULTO/ADULTO
      # on garde SUBADULTO
      any(Age == "SUBADULTO") ~ "SUBADULTO",
      # ADULTO sans ambiguïté
      any(Age == "ADULTO") ~ "ADULTO",
      
      TRUE ~ NA_character_ ),  .groups = "drop"  )

# is there an individual with different values in the same month ?
age_monthly %>%
  group_by(ID, Year, Month) %>%
  summarise( n_age = n_distinct(Age),  ages = paste(unique(Age), collapse = ", "), .groups = "drop" ) %>%
  filter(n_age > 1)


prop_age_monthly <- age_monthly %>%
  filter(!is.na(Age)) %>%
  mutate( Season = case_when( Month %in% 2:7 ~ "Wet", Month %in% c(8:12, 1) ~ "Dry" ) ) %>%
  count(Fragment, Species, Season, Age) %>%
  group_by(Fragment, Species, Season) %>%
  mutate( Proportion = n / sum(n)  ) %>%
  ungroup() 

final_prop_age <- prop_age_monthly %>%
dplyr::select(-n) %>% mutate(Parameter = str_replace_all(paste("Prop", Age, Season), " ", "_"))%>% dplyr::select(-c(Season, Age)) %>%  pivot_wider(names_from = Fragment, values_from = Proportion) %>% dplyr::select(Species, everything(), Parameter)

# Visualize this


ggplot( prop_age_monthly %>% filter(Species == "Didelphis_albiventris"),
  aes(x = Season, y = Proportion, fill = Age)) +
  geom_col(position = position_dodge()  ) +
  facet_wrap(~Fragment) +
  theme_bw() +
  labs( x = "Season", y = "Proportion",  fill = "Age")


ggplot( prop_age_monthly %>% filter(Species == "Marmosa_murina"),
  aes(x = Season, y = Proportion, fill = Age)) +
  geom_col(position = position_dodge()  ) +
  facet_wrap(~Fragment) +
  theme_bw() +
  labs( x = "Season", y = "Proportion",  fill = "Age")


```


## Final table 

```{r}

rbind(Sex_ratio, final_percent_repro, final_prop_age)

```



